OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-03-16 paper/paper.Rmd
Add TO-DOs from mtg w/ Pat

We performed _de novo_ clustering with OptiClust at distance threshold of 4%
using the Greengenes database, SILVA non-redundant database, or the Ribosomal
Database Project (RDP) [@desantis_greengenes_2006; @quast_silva_2013;
@cole_ribosomal_2014].
We then performed refernce clustering with OptiFit, fitting each dataset to each 
set of database OTUs repeated 100 times.
We measured the algorithm run time and evaluated OTU quality using the MCC as
described previously [@westcott_opticlust_2017].
To compare to existing software, we also clustered each dataset into _de novo_
OTUs  using OptiClust and VSEARCH, and with VSEARCH in reference-based mode
against the Greengenes database.

@@

@@

OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode (`r mcc_opticlust_vs_fit_split`% difference in median MCC).
OTU quality was remarkably stable across reference fraction sizes within
parameter sets (Fig. **XX**).
For example, splitting the human dataset 100 times yielded a coefficient of
variation of `r cv_fit_split_mcc_human` for the MCC score.
In terms of runtime, closed and open reference OptiFit performed faster than
OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and 
`r sec_open_fit_split_vs_clust` respectively.
The split dataset strategy performed `r sec_closed_fit_split_vs_db`% faster than
the database strategy in closed reference mode and `r sec_open_fit_split_vs_db`%
in open reference mode.

We also tested three methods for selecting the fraction of sequences to be used
as the reference; a simple random sample, weighting sequences by relative
abundance, and weighting by similarity to other sequences in the dataset.
OTU quality was similar with the simple and abundance-weighted sampling (median
MCCs `r mcc_fit_split_simple` and `r mcc_fit_split_abun` respectively), but
worse for similarity-weighted sampling with a median MCC of `r mcc_fit_split_dist`.
In closed reference mode, the fraction of query sequences that can be fit to the
reference OTUs increases as the reference fraction increases; from 
`r frac_fit_split_0.1`% of query sequences fit when using 10% of the dataset as 
the reference, to `r frac_fit_split_0.8`% of query sequences fit when using 80% 
of the dataset as the reference.

>>>>

- TODO: start paragraph with question, end with why we should care. transitions to move the story along.
- start:"In order to compare Optifit.."
- end: "optifit performs better"
- AND, BUT, THEREFORE, structure
- order of importance: mcc, mapping, sec

**TODO: lead with comparison of optifit vs current state of the art. give context at beginning.**
We performed _de novo_ clustering with OptiClust at distance threshold of 3% (TODO R CODE HERE)
using the Greengenes database, SILVA non-redundant database, or the Ribosomal
Database Project (RDP) [@desantis_greengenes_2006; @quast_silva_2013;
@cole_ribosomal_2014].
We then performed refernce clustering with OptiFit, fitting each dataset to each 
set of database OTUs repeated 100 times.
We measured the algorithm run time and evaluated OTU quality using the MCC as
described previously [@westcott_opticlust_2017].
To compare to existing software, we also clustered each dataset into _de novo_
OTUs  using OptiClust and VSEARCH, and with VSEARCH in reference-based mode
against the Greengenes database.

since closed ref threw out a lot of stuff, we performed open ref.

@@

- TODO; start with problem with current state of the art. 
- tied to reference dbs taht may be unrelated to samples studying. can we use our own data as a reference?
- lay out the use case and how this is a benefit over ref-db clustering.
- we wanted ot use x% of data as ref, weren't sure what size to use and strategy for selecting.
- important result: it works!

@@

TODO; ref weight simple only here.
OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode (`r mcc_opticlust_vs_fit_split`% difference in median MCC).
OTU quality was remarkably stable across reference fraction sizes within
parameter sets (Fig. **XX**).
For example, splitting the human dataset 100 times yielded a coefficient of
variation of `r cv_fit_split_mcc_human` for the MCC score.
In terms of runtime, closed and open reference OptiFit performed faster than
OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and 
`r sec_open_fit_split_vs_clust` respectively.
The split dataset strategy performed `r sec_closed_fit_split_vs_db`% faster than
the database strategy in closed reference mode and `r sec_open_fit_split_vs_db`%
in open reference mode.
TODO: fraction mapped.

TODO: pick a fraction (e.g. 50%). this comparison would be supplemental if there's a figure.
We also tested three methods for selecting the fraction of sequences to be used
as the reference; a simple random sample, weighting sequences by relative
abundance, and weighting by similarity to other sequences in the dataset.
OTU quality was similar with the simple and abundance-weighted sampling (median
MCCs `r mcc_fit_split_simple` and `r mcc_fit_split_abun` respectively), but
worse for similarity-weighted sampling with a median MCC of `r mcc_fit_split_dist`.
In closed reference mode, the fraction of query sequences that can be fit to the
reference OTUs increases as the reference fraction increases; from 
`r frac_fit_split_0.1`% of query sequences fit when using 10% of the dataset as 
the reference, to `r frac_fit_split_0.8`% of query sequences fit when using 80% 
of the dataset as the reference.
